---
title: "Kidney Exchange Pools"
author: "Nick Arnosti"
date: "2024-03-31"
output:
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{tikz}
- \usetikzlibrary{arrows.meta}
- \usepackage{array}
draft: true
---



Monthly batching. At the end of each month, you have a directed graph. Maximize total matches. 


<!--Let $D$ be the graph representing who can give to who. That is $D[i,j] = 1$ if and only if -->

Let $S$ be the set of patient-donor pairs. 

Let $C \subset S\times S$ be the set of compatible pairs. That is $(i,j) \in C$ means that the donor from pair $i$ can give to the patient from pair $j$. $\color{red}{\text{Represent with compatibility graph?}}$

We want to some up with a transplant plan $T \subset C$. It will be convenient to let this plan be represented by a set of variables $x[i][j] \in {0,1}$, where $x[i,j] = 1$ means that $(i,j) \in T$ (that is, $i$ donates to $j$) and $x[i,j] = 0$ means that $(i,j) \not\in T$.

\begin{align}
\max \sum_{i,j} x[i,j] & \text{Maximize total transplants} \\
\hspace{0.5 in}  x[i,j] \le C[i,j]  & \forall i,j \in S  \hspace{.5 in} \text{Transplants can occur only between compatible patients and donors} \\
\hspace{0.5 in} \sum_{j} x[i,j] \le 1  & \forall i \in S  \hspace{.5 in} \text{Each donor can give to at most one recipient} \\
\hspace{0.5 in} \sum_{j} x[j,i] \le 1  & \forall i \in S  \hspace{.5 in}  \text{Each patient can receive from at most one donor} \\
\hspace{0.5 in} \sum_{j} x[i,j] \le \sum_{j} x[j,i] & \forall i \in S \hspace{.5 in} \text{Each donor gives only if their patient receives} \\
\end{align}

This is an integer programming problem with linear constraints and objectives. $\color{blue}{\text{Natural question: if we relax integrality constraint, is it naturally integral? Intuitively and based on simulations, yes. But total unimodularity is a total mystery to me.}}$

$\color{red}{\text{Construct bipartite graph.}}$

Given the directed graph $G$, construct the following (undirected) bipartite graph $(D,P)$, where $D$ represents the set of donors and $P$ represents the set of patients. 

We have the following neat observation: if $M$ is a perfect matching in the bipartite graph $(D,P)$ (meaning that every vertex in $M$ is matched), then $M$ corresponds to a set of disjoint cycles in $G$. Let's do an example. 

$\color{red}{\text{Solve maximum weight bipartite matching (i.e. with Hungarian algorithm -- can go into that).}}$


In practice, it's illegal to write a contract involving the donation of a kidney. Thus, if patient A receives a kidney before donor A gives one, donor A could in theory back out of any planned donation with no ramifications. For this reason, cycles are generally completed using simultaneous surgeries. A two-way exchange requires four surgical teams and operating rooms. A three-way exchange requires six surgical teams. For this reason, most exchanges limit the length of cycle that can be used. 

Let's start with the case of 2-way cycles. We can modify the program above to add the following constraint:
\begin{equation} x[i,j] = x[j,i] \end{equation}
This enforces that we can only do 2-cycles by requiring that if $i$ donates to $j$, then $j$ donates to $i$.

Now, the relaxation into a linear program is definitely not integral. Consider the following example, with three pairs. Any donor can donate to both of the other patients. Because there are only three pairs and we are restricted to two-way exchanges, the best we can do is match two of the three pairs. 
```{r tikz-edges, results='asis', echo=FALSE}
# Define adjacency matrices M (perfect matching) and G (all edges)
generate_random_perfect_matching_matrix <- function(n) {
  repeat {
    permutation <- sample(n)
    if (!any(permutation == 1:n)) {
      M <- matrix(0, nrow = n, ncol = n)
      M[cbind(1:n, permutation)] <- 1
      return(M)
    }
  }
}

set_non_diagonal_zeros_to_ones <- function(M, k) {
  non_diagonal_zeros <- which(M == 0 & row(M) != col(M), arr.ind = TRUE)
  
  if (nrow(non_diagonal_zeros) < k) {
    stop("Not enough non-diagonal zeros to change.")
  }
  
  selected_indices <- sample(nrow(non_diagonal_zeros), k)
  M[non_diagonal_zeros[selected_indices,]] <- 1
  
  return(M)
}

generate_bipartite_graph_tikz_figure <- function(G, M = 0*G) {
  n <- nrow(G) # Assuming M and G are square and of the same size
  cat("\\begin{tikzpicture}[vertex/.style={circle,draw,minimum size=1.5em}, matched/.style={line width=2pt}, unmatched/.style={}]\n")
  
  for (i in 1:n) {
    cat(sprintf("\\node[vertex] (D%d) at (2*%d,3) {D%d};\n", i, i, i))
    cat(sprintf("\\node[vertex] (P%d) at (2*%d,0) {P%d};\n", i, i, i))
  }
  
  matched_edges <- which(M == 1, arr.ind = TRUE)
  if (length(matched_edges) > 0) {
    for (idx in 1:nrow(matched_edges)) {
      cat(sprintf("\\draw[matched] (D%d) -- (P%d);\n", matched_edges[idx, "row"], matched_edges[idx, "col"]))
    }
  }
  
  unmatched_edges <- which(G == 1 & M == 0, arr.ind = TRUE)
  if (length(unmatched_edges) > 0) {
    for (idx in 1:nrow(unmatched_edges)) {
      cat(sprintf("\\draw[unmatched] (D%d) -- (P%d);\n", unmatched_edges[idx, "row"], unmatched_edges[idx, "col"]))
    }
  }
  
  cat("\\end{tikzpicture}\n\n")
}


generate_directed_graph_tikz_figure <- function(G,M=G*0) {
  n <- nrow(G) # Assuming M and G are square and of the same size
  radius <- 3  # Radius of the circle on which vertices are placed
  cat("\\begin{tikzpicture}[vertex/.style={circle,draw,minimum size=1.5em}, matched/.style={-Stealth, line width=2pt}, unmatched/.style={-Stealth}]\n")
  
  # Calculate positions and draw vertices
  for (i in 1:n) {
    angle <- 90 - 360 * (i - 1) / n
    x <- radius * cos(angle * pi / 180)
    y <- radius * sin(angle * pi / 180)
    cat(sprintf("\\node[vertex] (V%d) at (%f,%f) {V%d};\n", i, x, y, i))
  }
  
  # Draw matched and unmatched edges
  matched_edges <- which(M == 1, arr.ind = TRUE)
  if(length(matched_edges)>0){
    for (idx in 1:nrow(matched_edges)) {
      cat(sprintf("\\draw[matched] (V%d) -- (V%d);\n", matched_edges[idx, "row"], matched_edges[idx, "col"]))
    }
  }

  
  unmatched_edges <- which(G == 1 & M == 0, arr.ind = TRUE)
  if(length(unmatched_edges)>0){
    for (idx in 1:nrow(unmatched_edges)) {
      cat(sprintf("\\draw[unmatched] (V%d) -- (V%d);\n", unmatched_edges[idx, "row"], unmatched_edges[idx, "col"]))
    }
  }
  
  cat("\\end{tikzpicture}\n\n")
}


generate_constraint_matrix <- function(G) {
  n <- nrow(G)  # Number of vertices
  edges <- which(G == 1, arr.ind = TRUE)
  num_edges <- nrow(edges)
  
  # Initialize the constraint matrix A and RHS vector b
  # 2n constraints for incoming and outgoing, n constraints for balance
  A <- matrix(0, nrow = 3 * n, ncol = num_edges)
  b <- c(rep(1, 2 * n), rep(0, n))  # Outgoing, Incoming constraints <= 1, balance constraints <= 0
  
  # Labeling columns (variables x_ij)
  #col_labels <- paste0("x[", edges[, "row"], ",", edges[, "col"], "]")
  col_labels <- paste0("x", edges[, "row"], edges[, "col"])

  # Setting up the constraints
  for (vertex in 1:n) {
    # Outgoing edge constraints for vertex
    outgoing <- which(edges[, "row"] == vertex)
    A[vertex, outgoing] <- 1  # Each vertex can have at most one outgoing edge
    
    # Incoming edge constraints for vertex
    incoming <- which(edges[, "col"] == vertex)
    A[n + vertex, incoming] <- 1  # Each vertex can have at most one incoming edge
    
    # Balance constraints: Outgoing - Incoming <= 0
    A[2 * n + vertex, outgoing] <- 1   # Add 1 for outgoing edges
    A[2 * n + vertex, incoming] <- -1  # Subtract 1 for incoming edges
  }
  
  # Generate row labels
  outgoing_labels <- paste("Donate", 1:n)
  incoming_labels <- paste("Receive", 1:n)
  balance_labels <- paste("IR", 1:n)
  row_labels <- c(outgoing_labels, incoming_labels, balance_labels)
  
  list(A = A, b = b, col_labels = col_labels, row_labels = row_labels)
}

#By default, assumes cycles of unbounded length, and no altruistic donors
generate_constraint_matrix <- function(G, altruistic = c(), pairwise_only = FALSE) {
  n <- nrow(G)  # Number of vertices
  edges <- which(G == 1, arr.ind = TRUE)
  num_edges <- nrow(edges)
  
  # Calculate the total number of constraints
  total_constraints <- 2 * n + ifelse(pairwise_only, num_edges, n - length(altruistic))
  
  # Initialize the constraint matrix A, RHS vector b, and inequalities vector
  A <- matrix(0, nrow = total_constraints, ncol = num_edges)
  b <- c(rep(1, 2 * n), rep(0, total_constraints - 2 * n))  # Adjust b based on the number of constraints
  inequalities <- c(rep("$\\leq$", total_constraints))  # Default inequality
  
  # Labeling columns (variables x_ij)
  col_labels <- paste0("x", edges[, "row"], edges[, "col"])

  # Setting up the constraints
  # Outgoing and Incoming Edge Constraints
  for (vertex in 1:n) {
    # Outgoing edge constraints for vertex
    outgoing <- which(edges[, "row"] == vertex)
    A[vertex, outgoing] <- 1
    
    # Incoming edge constraints for vertex
    incoming <- which(edges[, "col"] == vertex)
    A[n + vertex, incoming] <- 1
  }
  
  # Balance Constraints (excluding altruistic vertices)
  balance_constraints_start <- 2 * n + 1
  for (vertex in setdiff(1:n, altruistic)) {
    outgoing <- which(edges[, "row"] == vertex)
    incoming <- which(edges[, "col"] == vertex)
    A[balance_constraints_start, outgoing] <- 1
    A[balance_constraints_start, incoming] <- -1
    balance_constraints_start <- balance_constraints_start + 1
  }
  
  # Pairwise Constraints (if applicable)
  # This could not work well if G[i,j] \neq G[j,i], as we only create variables for edges present in the graph. Thus, expressing the pairwise constraint could be difficult.
  if (pairwise_only) {
    pairwise_constraints_start <- balance_constraints_start
    for (e in 1:num_edges) {
      A[pairwise_constraints_start, e] <- 1
      corresponding_edge <- which(edges[, "row"] == edges[e, "col"] & edges[, "col"] == edges[e, "row"])
      if (length(corresponding_edge) == 1) {
        A[pairwise_constraints_start, corresponding_edge] <- -1
        #inequalities[pairwise_constraints_start] <- "="  # Set equality for pairwise constraints
      }
      pairwise_constraints_start <- pairwise_constraints_start + 1
    }
  }
  
  # Generate row labels
  outgoing_labels <- paste("Donate", 1:n)
  incoming_labels <- paste("Receive", 1:n)
  balance_labels <- setdiff(paste("IR", 1:n), paste("IR", altruistic))
  pairwise_labels <- if (pairwise_only) paste("Pairwise", edges[, "row"], edges[, "col"]) else character()
  row_labels <- c(outgoing_labels, incoming_labels, balance_labels, pairwise_labels)
  
  list(A = A, inequalities = inequalities,b = b, col_labels = col_labels, row_labels = row_labels) 
}


generate_constraint_matrix <- function(G, altruistic = c(), pairwise_only = FALSE) {
    n <- nrow(G)  # Number of vertices
    edges <- which(G == 1, arr.ind = TRUE)
    num_edges <- nrow(edges)
    
    # Initial count of constraints includes outgoing, incoming, and balance for each vertex
    num_constraints <- 2 * n + n - length(altruistic)
    
    # Adjust for additional pairwise constraints
    additional_constraints <- 0
    if (pairwise_only) {
        for (e in 1:num_edges) {
            i <- edges[e, "row"]
            j <- edges[e, "col"]
            # Add a constraint for unidirectional edges or for bidirectional edges with i < j to avoid duplication
            if (G[i,j] == 1 && G[j,i] == 0 || (G[i,j] == 1 && G[j,i] == 1 && i < j)) {
                additional_constraints <- additional_constraints + 1
            }
        }
    }
    # Total constraints now also includes the additional pairwise constraints
    num_constraints <- num_constraints + additional_constraints
    
    A <- matrix(0, nrow = num_constraints, ncol = num_edges)
    b <- c(rep(1, 2 * n), rep(0, n - length(altruistic) + additional_constraints))
    inequalities <- rep("$\\leq$", num_constraints)
    col_labels <- paste0("x", edges[, "row"], edges[, "col"])
    row_labels <- vector("character", num_constraints)
    
    # Populate row labels for outgoing, incoming, and balance constraints
    row_labels[1:n] <- paste("Donate", 1:n)
    row_labels[(n + 1):(2 * n)] <- paste("Receive", 1:n)
    balance_start <- 2 * n + 1
    balance_end <- balance_start + n - length(altruistic) - 1
    row_labels[balance_start:balance_end] <- paste("IR", setdiff(1:n, altruistic))
    
    # Setup matrix A for outgoing and incoming constraints
    for (vertex in 1:n) {
        outgoing <- which(edges[, "row"] == vertex)
        A[vertex, outgoing] <- 1  # Outgoing
        
        incoming <- which(edges[, "col"] == vertex)
        A[n + vertex, incoming] <- 1  # Incoming
    }
    
    # Setup matrix A for balance constraints, excluding altruistic vertices
    balance_index <- 2 * n + 1
    for (vertex in setdiff(1:n, altruistic)) {
        outgoing <- which(edges[, "row"] == vertex)
        incoming <- which(edges[, "col"] == vertex)
        A[balance_index, outgoing] <- 1
        A[balance_index, incoming] <- -1
        balance_index <- balance_index + 1
    }
    
    # Setup matrix A for pairwise constraints
    if (pairwise_only) {
        pairwise_index <- balance_index
        for (e in 1:num_edges) {
            i <- edges[e, "row"]
            j <- edges[e, "col"]
            if (G[i,j] == 1 && G[j,i] == 0) {
                A[pairwise_index, e] <- 1
                b[pairwise_index] <- 0
                inequalities[pairwise_index] <- "$=$"
                row_labels[pairwise_index] <- paste("Pairwise", i, j)
                pairwise_index <- pairwise_index + 1
            } else if (G[i,j] == 1 && G[j,i] == 1 && i < j) {
                corresponding_edge_idx <- which(edges[, "row"] == j & edges[, "col"] == i)
                A[pairwise_index, e] <- 1
                A[pairwise_index, corresponding_edge_idx] <- -1
                b[pairwise_index] <- 0
                inequalities[pairwise_index] <- "$=$"
                row_labels[pairwise_index] <- paste("Pairwise", i, j)
                pairwise_index <- pairwise_index + 1
            }
        }
    }
    
    list(A = A, b = b, col_labels = col_labels, row_labels = row_labels, inequalities = inequalities)
}



print_constraints_with_rhs <- function(A, row_labels, col_labels, inequalities, b) {
  # Check if the length of inequalities vector matches the number of rows in A
  if (length(inequalities) != nrow(A)) {
    stop("The length of the inequalities vector must match the number of rows in A.")
  }
  
  column_width <- "1cm"  # Adjust this based on your content and preference
  rhs_column_width <- "1cm"  # Adjust as needed
  
  # Start the tabular environment with column specifications
  cat("\\begin{tabular}{r")  # 'r' for the first column (row names) to be right-aligned
  # Add specifications for other columns (variables)
  for (i in 1:length(col_labels)) {
    cat(" >{\\raggedleft\\arraybackslash}p{", column_width, "}")
  }
  # Add specifications for inequality and RHS columns, no header needed
  cat(">{\\raggedleft\\arraybackslash}p{2cm}")  # for the inequality column
  cat(">{\\raggedleft\\arraybackslash}p{", rhs_column_width, "}")  # for the RHS values
  cat("}\n")
  
  # Print the header row with column names (leave last two headers blank)
  cat("\\textbf{Constraint} & ", paste("\\textbf{", col_labels, "}", sep="", collapse=" & "), " & & \\\\\\hline\n")
  
  # Iterate over each row to print its contents along with inequalities and RHS values
  for (i in 1:nrow(A)) {
    # Prepare the content of the row by marking non-zero elements
    row_content <- sapply(1:ncol(A), function(j) {
      if (A[i, j] != 0) {
        return(as.character(A[i, j]))
      } else {
        return("")
      }
    })
    # Combine the row label, its content, inequality, and RHS value, then print
    cat(row_labels[i], " & ", paste(row_content, collapse=" & "), " & ", inequalities[i], " & ", b[i], " \\\\\n")
  }
  
  # End the tabular environment
  cat("\\end{tabular}\n\n")
}


G = rbind(c(0,0,1,0,1,0),c(1,0,0,0,0,0),c(0,1,0,1,0,0),c(0,0,0,0,1,0),c(0,0,1,0,0,1),c(1,0,0,0,0,0))
#Generating a drawing of the graph associated with G
generate_directed_graph_tikz_figure(G)

G = rbind(c(0,1,1),c(1,0,1),c(1,1,0))
#Generating a drawing of the graph associated with G
generate_directed_graph_tikz_figure(G)

# Printing the constraint matrix A with row and column names
result <- generate_constraint_matrix(G,pairwise_only=TRUE)
print_constraints_with_rhs(result$A, result$row_labels, result$col_labels, result$inequalities, result$b)
```




Hello World

```{r example, results='asis', echo=FALSE}
# Generate random graph and matching, then call the function to print the TikZ commands
set.seed(2024)
n = 7
k = 14
M <- generate_random_perfect_matching_matrix(n)
G <- set_non_diagonal_zeros_to_ones(M,k)
generate_directed_graph_tikz_figure(G)
generate_directed_graph_tikz_figure(G,M)
generate_bipartite_graph_tikz_figure(G,M)
```



Things to consider in optimization class:

* In general undirected graph, say that a set of agents is feasible if there is a matching that matches them all. The family of feasible sets of agents forms a matroid. Therefore (i) every Pareto efficient matching has equal size, (ii) We can use a greedy algorithm to find a max weight matching (still need to go into some details).
* Could define weight of an edge to be sum of weights of its two vertices, and then use a greedy algorithm to select edges. This will lead to a maximal matching (thus within 1/2 of maximum cardinality matching), and also should lead to within 1/2 of maximum weight matching.
* Matroids, greedy algorithms, linear programming, LP relaxations, hierarchy, bi-hierarchy, polynomial time, NP hardness and reductions, 




