---
title: 'Lung Allocation'
date: '2025-10-02'
author: Nick Arnosti

draft: true
nurmarkdownumber_sections: true
---

\usepackage{amsthm}
\newtheorem{conjecture}{Conjecture}
\newtheorem{definition}{Definition}

```{r}
library(data.table)
initial_patients = c()
arriving_patients = c(1,2,3)
donor_data = data.table(match_id = c(7,8,9,10))
patient_data = data.table(px_id=arriving_patients,n_accept=c(1,0,1),match_runs=c(4,0,2))
```

Let's look at what happened. 

In this time horizon, we had 
```{r} 
length(arriving_patients)
``` 
arriving patients and 
```{r} 
length(donor_data$match_id)
```
arriving donors.

Of our arriving patients, 
```{r} 
sum(patient_data$n_accept > 0) 
```
accepted at least one offer. 


# Unmatchable Agents

There are
```{r} 
unmatchable_patients = patient_data[match_runs==0,px_id] 
length(unmatchable_patients)
```
patients in the data set who never receive a sequence number. That is, these patients are screened out of every match_run that occurs while they are present. 

Of these, 
```{r} 
sum(unmatchable_patients %in% initial_patients)
```
are patients who were waiting at the beginning of the window.


# Measuring Organ Availability and Prioritizing Hard to Match Patients

It is generally acknowledged that some patients are harder to match, due to concerns such as blood type, tissue antibodies, and height. As a result, organ policies often give "priority" to certain groups to try to "level the playing field" and compensate for their disadvantage. For example, for lungs, patients with type $O$ blood are given 5 priority points, and for kidneys, patients with high CPRA are prioritized (how?).

The goal of this note is to examine whether these policies help to make the allocation more "fair." To answer this question, we must propose a definition of fairness. In addition, we propose a measure of organ availability.

We first consider a static model with a fixed population of patients $P$ and donors $D$. There is an edge from $p$ to $d$ if that match is feasible (this edge could consider both medical compatibility and temporal overlap). We denote the set of edges by $E$.

One natural measure of how many organs are available to a patient $p$ is the number of outgoing edges. (In a dynamic model, this would correspond to the rate at which compatible organs arrive to the market.) To see why this is a flawed metric, consider the following example.

**Example 1**. *There are $75$ patients with blood type $A$ and $25$ with blood type $B$. There are $65$ donors with blood type $A$ and $35$ with blood type $B$.*

In this example, the type $A$ patients are compatible with more organs than the type $B$ patients are, but the type $B$ patients should have no trouble receiving a transplant (there is a surplus of type $B$ donors) whereas some type $A$ patients will go without. In other words, organs will be more "available" to a type $B$ patient.

The problem with only looking at number of outgoing edges is that it does not consider the competition from other patients (i.e. how many patients are seeking the same organs). 

I propose the following measure of organ availability.

\begin{align}
\max \sum_{p \in P} m_p \log(q_p) \\
\sum_{d \in D} x_{pd} & = q_p \nonumber \\
\sum_{p \in P}m_p x_{pd} & \le n_d \nonumber \\
x_{pd} & \le {\bf 1}((p,d) \in E) \nonumber \\
q_p & \le 1 \nonumber
\end{align}
*Baseline, $m_p = n_d = 1$. But this also works if $p$ and $d$ represent types of patients and donors, and $m_p$ and $n_d$ represent the number of patients or donors of these types.*

::: {.theorem #label_name name="Theorem Name"}
Let $x^*$ be an optimal solution. Then there exist costs $c_d$ that support $x^*$ as a competitive equilibrium from equal incomes.
:::

There may be some patients with no compatible donors, for whom $q_p$ is necessarily zero. We can remove these patients from the objective function above before solving the optimization problem.

The quantity $q_p$ will be our measure of "availability" for patient $p$. In particular, for Example 1 with $m_A = 75$, $m_B = 25$, $n_A = 65$ and $n_B = 35$, we would have $q_A = 65/75$ and $q_B = 1$. More generally, $q_A < q_B$ will imply $m_A/n_A < m_B/n_B$.

Let's consider a different example.

**Example 1**. *There are $60$ patients with blood type $O$ and $40$ with blood type $A$. There are $60$ donors with blood type $O$ and $30$ with blood type $A$.*

In this example, note that if we naively compute the ratios $m_O/n_O$ and $m_A/n_A$, we see that the former is larger. So we might expect that organs are more available to $O$ patients. But of course this ignores that $O$ organs can also be given to $A$ patients. Since the set of organs available to $A$ patients is a superset of those available to $O$ patients, we might argue that organs are more available to $A$ patients. But if we solve the optimization problem above, we get $q_O = q_A = 0.9$, in an allocation where 6 $O$ organs are given to $A$ patients. More generally, if we change the numbers in Example 2, we will see that so long as $m_O/n_O \ge m_A/n_A$, $q_O$ will be equal to $q_A$. If $m_O/n_O < m_A/n_A$ and $m_O/n_O < 1$, then we will have $q_O < q_A$.


The following gives our definition of what it means for an allocation to be fairer than another.
::: {.definition #label_name name="Definition Name"}
Given $q \in [0,1]^P$, let $q_{(k)}$ denote the $k^{th}$ smallest value of $q$. We say that $q$ **Lorentz Dominates** $q'$ if for all $k \in \{1, \ldots, |P|\}$, we have $\sum_{j = 1}^k q_{(j)} \ge \sum_{j = 1}^k q'_{(j)}$.
:::



Note that this intuitively captures both fairness and efficiency: if $q$ Lorentz dominates $q'$, then $q$ must allocate at least as much in aggregate as $q'$. In particular, any Pareto improvement upon $q'$ will Lorentz dominate $q'$. In this context, I think this is appropriate: nobody is suggesting that we waste certain organs that could go to more advantaged patients in order to make the allocation more "fair". 

::: {.theorem #label_name name="Theorem Name"}
Let $x$ be feasible for \eqref{eq:}, and let $x^*$ be optimal for \eqref{eq:}. Then $q(x^*)$ Lorentz dominates $q(x)$.
:::
(Note that $x^*$ may not be unique, but $q(x^*)$ is unique.)

**Example 3** (Non-uniqueness). There are 60 $O$, 30 $A$, 10 $AB$ patients, and 60 $O$, 27 $A$ and 3 $AB$ donors. Then one optimal solution gives 6 $O$ organs to AB patients. However, any solution of the form $X_{O-O} = 54$, $X_{AB-AB} = 3$, $X_{AB-O} = k$, $X_{AB-O} = k$, $X_{AB-A} = 6-k$, $X_{A-O} = 6-k$, $X_{A-A} = 21+k$ is optimal. (90% of each group is matched.)


We can now do a calibrated exercise where we choose the arrival rates of patients and organs based on the data. If we include all arrivals, we simply get $q_T = 1$ for all blood types $T$. Therefore, we will include only arrivals that were ultimately accepted by some patient. In this case, we see very few $AB$ donors, but nevertheless, all agents have the same $q$. 


**Future Work** Incorporate agent preferences for being matched sooner? Look at outcomes in some sort of serial dictatorship (in organ arrival order) or DA?

Could we look at acceptance patterns (correlation between height of donor and height of patient)? Use to come up with a model of which height ranges can accept each other (based on horseshoe shape).


# Dynamic Matching: the surprising effectiveness of First Come First Served
 
Consider a dynamic model where paients of type $p$ arrive according to a Poisson process at rate $\lambda_p$ and abandon after a sojourn time that is drawn from distribution $F$. Donors of type $d$ arrive according to a Poisson process at rate $\lambda_d$, and must be matched immediately or discarded. There is a set of edges $E$ which indicates which donor types can be allocated to each patient type.

A first come first served processing rule always gives an organ to the patient who has waited longest, among those who are compatible with it.

For $p\in P$ and $d \in D$, define $A_p(t)$ to be the number of patients of type $p$ who arrive in the interval $[0,t]$ and define $N_{pd}(t)$ to be the number of transplants between a patient of type $p$ and a donor of type $d$ on this interval.  Define $x^{FCFS}_{pd}(t) = N_{pd}(t)/A_p(t)$.

::: {.lemma #label_name name="Lemma Name"}
For any arrival rates $\lambda$, edges $E$, and sojourn time distribution $F$, $x^{FCFS}_{pd}(t)$ converges almost surely to $\overline{x}^{FCFS}_{pd}$. *Of course, $\overline{x}$ depends on $\lambda$ and $E$, but I suspect that it does not depend on $F$.*
:::


::: {.theorem #label_name name="Theorem Name"}
Fix $E$. As $n \rightarrow \infty$, $\overline{x}^{FCFS}_{pd}(n \lambda)$ converges to an optimal solution to \eqref{eq:}.
:::

Note that this holds for any $\lambda$ and any $E$ and any $F$.

